An Ordered Logic for Category Theory

### 2022-09-07 paper/paper.tex
some work on the semantics section

\begin{abstract}
  We present a domain-specific type theory for constructions and
  proofs in category theory. The type theory axiomatizes notions of
  category, functor, profunctor and a generalized form of natural
  transformations. By restricting the type theory to an ordered
  linear restriction of higher-order logic, we guarantee that all
  mappings between categories are functorial and all transformations
  are natural by construction, with no separate proof
  necessary. Important category theoretic proofs such as the Yoneda
  lemma and Co-yoneda lemma become simple type theoretic proofs about
  the relationship between unit, tensor and (ordered) function types,
  and can be seen to be ordered refinements of theorems in
  intuitionistic higher-order logic.

  We give a sound and complete notion of model of our type theory in a
  structure we call a hyperdoctrine of pro-arrow equipments with
  displays, which are known to model enriched and internal category
  theory. This means that while the proofs in our type theory look
  like standard set-based arguments, the substructural syntactic
  restrictions ensure that the proofs carry over to enriched and
  internal settings as well.
\end{abstract}

@@

Category theory is a branch of mathematics that gives a systematic
theory of typed, ``higher-dimensional'' algebraic structures.
%
Originally developed for applications to homological algebra, it was
quickly discovered that these algebraic structures were common in
logic and computer science, and that formal systems like type theories
and programming languages can be modeled by a sound and complete
notion of semantic model given by some notion of structured
category\cite{lawvere,goguen,lambek,moggi}.
%
Examples include the simply typed lambda calculus \cite{lambekscott},
computational lambda calculus \cite{moggi}, linear logic \cite{girard}
and dependent type theory\cite{not-sure}.
%
These seminal works have led to a fruitful field of research applying
such categorical logic techniques to the design of new type theories
based directly on category-theoretic semantics.
%
The slogan is that the syntax of a type theory should present an
initial object in its category of models, a category-theoretic
reformulation and generalization of the logical notion of soundness
and completeness.

% TODO: do we have citations for this?
While this research program has been quite successful,
category-theoretic techniques in computer science are sometimes
criticized for being overly complex, and there is difficulty in
acquisition of competence in category theory by computer scientists
who do not have a strong background in ``pure'' mathemantics.
%
While much of this can be explained by lack of, or lack of awareness
of, accessible material for computer scientists, in some ways category
theory is objectively complex.
%
In a traditional set-theoretic formulation, notions such as adjoint
functors and Kan extensions produce a proliferation of ``naturality''
and ``functoriality'' side-conditions that must be discharged.
%
This is compounded when we move from ordinary category theory to
enriched and internal category theory, where constructions must be
additionally proven continuous, monotone, etc, in addition to natural
or functorial.
%
These generalizations have long been recognized as being important in
programming language applications: domain-,metric- and
step-index-enriched categories have been used to model general
recursion and internal categories have been used to model
parametricity and gradual
typing\cite{order-enriched-cats,topos-of-trees,reflexive-graphs,double-cats-gradual-typing}.

@@

Fortunately, the tools of category theory itself can be employed to
simplify these constructions, specifically the tools of \emph{higher}
category theory.
%
Consider the analogous situation in differential calculus. When a
physicist writes down a function, they do not expand out the
$\epsilon-\delta$ definition of continuity and prove that it is
continuous. Rather, they are aware of certain \emph{syntactic
principles}, most importantly that composition of continuous functions
is continuous and that previously defined functions are known to be
continuous.
%
Similar principles apply to category theory itself: a composition of
functors is functorial, a composition of natural transformations is
natural, etc.
%
In the case of category theory, these principles can be formalized
using algebraic structures such as 2-categories, bicategories,
(virtual) double categories and pro-arrow equipments\cite{}.
%
This approach has the advantage of avoiding the need to prove
the side conditions of naturality or enrichment.
%
The downside is that these algebraic structures are quite complicated,
with a 2-dimensional diagrammatic language that can be quite beautiful
and elegant, but is also seems very far removed from the traditional
formulation of category theory that while complex is at least
described in terms of a concrete notion of sets or types.

In this work we show that these algebraic formulations of category
theory can be given a type-theoretic syntax that allows us to make
``na\"ive'' set theoretic constructions for categories, but by
adhering to a certain syntactic discipline provide correct arguments
for not just ordinary category theory but enriched and internal
categories as well.
%
The type theory, which we call SCT, is a ``synthetic'' theory of
categories, functors, profunctors and natural transformations.
%
This provides a formulation of category theory in which profunctors
play a primary role, most recognizable in ordinary category theory in
the prominence of ``hom isomorphism'' formulations of universal
properties.
%
The semantics of SCT is given by a slight generalization of virtual
double categories \cite{} that we call double categories with
displays.
%
Additionally, to define signatures and parameterize theorems like the
Yoneda lemma, SCT includes an ordinary dependent type theory over it
which includes usual $\Pi,\Sigma,\Id$ types.
%
In an implementation of SCT, this dependent type theory might be the
host language itself.

In this work, we focus on the syntax and semantics of SCT, but the
ultimate goal is for this theory to serve as the foundation for a
framework in which we can more easily study and program with new type
theories. This falls in the tradition of \emph{logical frameworks}
such as the LF system. To provide evidence for the feasibility of this
approach, we show that we can model structures used in recent work on
modal dependent type theory. This shows that an implementation of SCT
would provide at least a toy prototype implementation/type checker for
modal dependent type theory. This can be summarized in the slogan
\begin{quote}
  category theory as a logical framework
\end{quote}
that is, that a syntax for category theory can be used as a logical
framework for type theories with semantics in structured categories.
%
If implemented in a proof assistant this would also provide with it
the soundness of the syntax in any model.
%
The emphasis on profunctors in SCT fits nicely with the
\emph{judgmental} approach of Per-Martin L\"of and suggests the
following slogan
\begin{quote}
  profunctors as judgments, natural transformations as derivations
\end{quote}
because we observe that the ``derivation tree'' style of writing
judgments corresponds closely to the type structure of profunctors and
natural transformations in SCT, and it should be simple to implement a
program that converts formal natural transformations into derivation
tree syntax.

This work is merely a first step in this direction, as there are still
three major components of category theory: opposite categories,
functor categories and ``universes'' such as the category of sets or
categories that are not nicely axiomatizable in the current system.
%
Nontheless, a great deal of category theory can be carried out in this
more limited setting, and we can still take advantage of duality at a
meta level.

\subsection{Ordered Logic of Functions and (Constructive) Relations}

Our logic is based on four basic notions: categories, functors,
profunctors and (multi-argument) natural transformations.
%
These generalize the basic notions already present in formulations of
higher-order logic: sorts, functions, relations and proofs.
%
In fact, one way to view our logic is that it is an \emph{ordered},
\emph{constructive} higher-order logic

While profunctors are not typically employed in introductory category
theory texts,

\subsection{Category theory with Profunctors}

First, we review the 4 basic notions of our theory of categories in
the traditional formulation. A category $\cat C$ is a set of objects
$\cat C_0$ with a span of arrows $\cat C_0 \leftarrow \cat C_1
\rightarrow \cat C_0$ that pick out the source and target of each
arrow, such that there are composition and identity arrows with
appropriate laws. A functor $F : \cat C \to \cat D$ consists of a
function on objects $F_0 : \cat C_0 \to \cat D_0$ and a function on
morphisms $F_1 : \cat C_1 \to \cat D_1$ that is compatible with source
and target and preserves identity and composition.

A profunctor $R$ from $\cat C$ to $\cat D$ is a span $\cat C_0
\leftarrow R \rightarrow \cat D_0$ that is compatible with
\emph{left}-composition by $\cat C$ and \emph{right}-composition by
$\cat D$. If we think of categories as generalizations of (ordered)
sets then functors generalize (monotone) functions and profunctors
generalize (monotone) \emph{relations}. We think of an element $r \in
R$ with $s(r) = c$ and $t(r) = d$ as a ``proof'' that $c R d$ or some
kind of ``heteromorphism'' from $c$ to $d$. The compatibility with
composition says that these relations are down-closed in $C$ and
up-closed in $D$, or that heteromorphisms can be composed with
homomorphisms.
%

%% Embracing the analogy ``profunctors as judgments'' we can view
%% elements of a profunctor $r \in R$ with $s(r) = A$ and $t(r) = B$ in
%% type-theoretic notation
%% \[ \inferrule*[right={r}]{~}{A \vdash_R B}\]
%% Noting that every category $\cat C$ has a hom profunctor $Hom_C : \cat
%% C \pto \cat D$, we get for each category a judgment
%% \[ \inferrule*{}{A \vdash_{Hom_C} A'} \]
%% and the category operations provide us with rules for use of the assumption:
%% \[ \inferrule*[right=id]{}{A \vdash_{Hom_C} A} \]
%% and cut
%% \[ \inferrule*[right=cut]{A \vdash_{Hom_C} A' \and A' \vdash_{Hom_C} A''}{A \vdash_{Hom_C} A''} \]
%% satisfying unit and associativity equations.
%% %
%% An arbitrary profunctor $R$ supports a kind of ``bi-cut''
%% rule\footnote{alternatively, this can be defined as 2 cuts: left and
%% right, with an associativity law}:
%% \[ \inferrule*[right=bicut]{A \vdash_{Hom_C} A' \and A' \vdash_R B' \and B' \vdash_{Hom_D} B}{A \vdash_R B} \]
%% That satisfies similar bi-unit and bi-associativity laws.

\subsection{Double Categories and Pro-arrow Equipments}

We can abstract the algebraic structure of categories, functors,
profunctors, and natural transformations as a \emph{double category}.
%
A double category is a generalization of categories where we have a
collection of objects $C_0$, and \emph{two separate} notions of
morphism, which we call vertical $C_v$ and horizontal $C_h$ over the
same objects $C_0$ which each have their own identity and composition,
and finally a notion of 2-cell that is most easily described

\[ % https://tikzcd.yichuanshen.de/#N4Igdg9gJgpgziAXAbVABwnAlgFyxMJZABgBpiBdUkANwEMAbAVxiRAGEB9YkAX1PSZc+QigCM5KrUYs2XMXwEgM2PASIAmSdXrNWiDpw2LBqkUQDM26XrYAdOwygQcCfqeHqUAFmu7ZBlzAYADUYrwmykJqoiSkYlL++iCRKp6xvgk6Msl8UjBQAObwRKAAZgBOEAC2SBIgOBBIWjYBIABKnAruIJU1zdSNSFatyZ3GPX21iCNDiL6jbJ2Ek1XTZA1NiPVJS9yRU0gbcwCs2bYGAGIHa0hnm0gAbOdtAMo3-fODW8+LBgDieV4QA
\begin{tikzcd}
C_0 \arrow[r, "R_0"] \arrow[d, "F"] & C_1 \arrow[r, "R_1"] & C_2 \arrow[r, "R_2"] & \ldots \arrow[r, "R_n"] & C_{n+1} \arrow[d, "G"] \\
{} \arrow[rrrr, "S"]                &                      &                      &                         & {}                    
\end{tikzcd} \]

This 2-cell has one output pro-arrow, many input pro-arrows which must
match the source from one to the target object of the next and a left
and right ``frame'' arrow. These can be composed in a manner similar
to multi-categories, where you can compose this with many input
2-cells, one for each input pro-arrow.
%
These then have complex associativity and unit rules and side
conditions that say the framing arrows compose alongside the 2-cells.

We address both of these issues by developing an alternative
\emph{type theoretic syntax} for virtual equipments.
%
In this syntax the ``multi-2-cell'' above instead becomes a natural
transformation judgment
\[ \Phi \vdash t : S[F(x);G(y)] \]
where the composable string of input arrows is modeled by an ordered,
linear context $\Phi$ and output $S$ and frames $F,G$ are combined in
the output type using the restriction operation of an equipment.
%
We write natural transformations in a natural deduction style: a
transformation is written as a ``parameterized'' element of the output
profunctor. Furthermore the restriction operators for vertical arrows
into horizontal arrows is naturally modeled by a syntactic
substitution.
%
This interpretation of restriction as substitution both matches the
semantic intuition for the restriction operation as well as greatly
simplifies reasoning about composition of transformations.

%% Experienced category theorists have no trouble with this, as they are
%% familiar with a great many principles for constructing transformations
%% that are guaranteed to be natural. Consider an analogy with
%% calculus. When a mathematician writes a formula such as

%% $$\sin(f(x) - g(x))$$

%% They do not need to spend time with an explicit $\epsilon,\delta$
%% proof that the function is continuous or furthermore smooth. Instead,
%% they follow syntactic principles that compositions of continuous
%% functions are continuous.

%% There are similar syntactic notions for guaranteeing that
%% transformations are natural.

%% \subsection{Category Theory with Profunctors}

%% The syntax of Judgment Theory gives not just a syntax for the familiar
%% trio of category, functor and natural transformation, but additionally
%% a concept that is not usually highlighted in introductory category
%% theory texts: the profunctor. A simple intuition is that profunctors
%% are to functors as relations are to functions. Indeed, the syntax for
%% judgment theory has a straightforward semantics in the category of
%% sets and relations.

%% The reason profunctors are not commonly introduced is that they are
%% seen as an auxiliary notion: a profunctor $R : C \pto D$ is
%% typically defined as a functor $C^{\textrm{op}}\times D \to
%% \textrm{Set}$. However this definition requires two concepts that are
%% somewhat problematic: the opposite category $C^\textrm{op}$ and the
%% category of (small) sets $\textrm{Set}$. However, these notions do not
%% necessarily exist in all of our intended applications. First, we are
%% interested in semantics where are objects are \emph{multicategories}
%% where morphisms have multiple inputs but only a single output. In this
%% case the opposite of a multicategory is not a multicategory at all but
%% instead some kind of ``co-multicategory'' with multiple outputs but
%% one input. Second, having a universe object like $\textrm{Set}$ is also

%% \subsection{Applications to Enriched, Internal and Multicategory Theory}

%% The syntax of judgment theory is an internal language for
%% \emph{virtual equipments}. These are a kind of joint generalization of
%% double categories and multicategories.

%% Notably, the definition of a category in \cref{sec:internal} can be
%% applied as an extremely general notion of category, including internal
%% categories and enriched categories. This general notion is often
%% called a ``generalized multi-category'' and encompasses an astonishing
%% range of mathematical structures, including categories, topological
%% spaces, metric spaces, etc.

%% \subsection{Applications to Generalized Type Theory}

%% While we mostly focus on the semantics in category theory, due to the
%% tight connection between category theory and the semantics of type
%% theory, there are direct applications to the study of ``abstract type
%% theory'' as well. We demonstrate how we can model single-variable
%% simple type theories with products and coproducts within SCT. This is
%% very much in the spirit of logical frameworks \cite{TODO}, since we
%% can faithfully model the terms of an object theory inside of our
%% metatheory. Since we have models of our metatheory in virtual double
%% categories, we establish sound semantics as well. Put succinctly,
%% category theory itself is the logical framework for the study of type
%% theories.

@@

\item In \cref{sec:examples} we show how to use our syntax for
  synthetic category theory. Much of this is just rendering well-known
  2-categorical/double categorical constructions in our type theoretic
  syntax to demonstrate the ``naive''/synthetic character of proofs
  and constructions in SCT. We additionally give some examples of how
  this can be applied to model a recent formulation of modal dependent
  type theory.

@@

\item In \cref{sec:semantics}, we show that the syntax of SCT is sound
  for models in an arbitrary virtual double category with
  corresponding universal constructions.

@@

We have aimed that the syntax and examples sections should be
accessible to anyone familiar with basic category theory and
willingness to understand profunctors and (co-)ends.
%
The semantics section is by necessity more advanced and may be skipped
or skimmed by readers who are willing to trust the mathematical
foundations of the system.

In Figure~\ref{fig:hol-cats} we give a table summarizing the
relationship between the judgments and connectives of higher-order
logic with our category logic. Since category logic is an ordered
variant of higher-order logic, some notions splinter into
multiple. For instance, covariant and contravariant presheaf
categories generalize the power set. Further, because we only have
binary relations rather than relations of arbitrary arity, we have
only restricted forms of universal and existential quantification
which combined with implications and conjunctions.

@@

@@

\subsection{Duality Principle}
One common source of category theoretic arguments is the duality
principle: since the opposite of any category is also a category, any
statement that quantifies over all categories is equally true of their
opposites. Furthermore in a formal language of categories, any
statement can be dualized to a statement where domains and codomains
are swapped and composition is flipped.
%
A similar duality principle holds in SCT.
%
The basic idea is to reverse the orientation of all \emph{set}
expressions by swapping which variable is covariant and which is
contravariant.
%
This has the effect of reversing the order of the context $\Phi$.
%
We can define this as a translation $\cdot^* : SCT \to SCT$ defined on
syntactic forms of SCT in Figure~\ref{fig:duality}.
\begin{figure}
    \begin{align*}
      (\harr P Q)^* &= \harr {P^*} {Q^*}\\
      (\alpha:\cat C)^* &= \alpha: \cat C^*\\
      (\alpha:\cat C; \beta:\cat D)^* &= \beta:\cat D^*; \alpha:\cat C^*\\
      (\Phi,x:R,\Psi)^* &= \Psi^*,x:R^*, \Phi^*\\
      (\homrallXYtoZ \alpha P Q)^* &= \homlallXYtoZ \alpha {P^*} {Q^*}\\
      (\homlallXYtoZ \alpha P Q)^* &= \homrallXYtoZ \alpha {P^*} {Q^*}\\
      (\homrlambdaXatYdotZ x \alpha s)^* &= \homllambdaXatYdotZ x \alpha s^*\\
      (\homrappXtoYatZ s t a)^* &= \homlappXtoYatZ {s^*} {t^*} {a^*}\\
      (\tensorexistsXwithYandZ \beta P Q)^* &=
      \tensorexistsXwithYandZ \beta {Q^*}{P^*}\\
      (\tensorintroatXwithYandZ b s t)^* &= \tensorintroatXwithYandZ {b^*} {t^*} {s^*}\\
      (\punitinXfromYtoZ {\cat C} {a_1}{a_2})^* &= \punitinXfromYtoZ {\cat C^*} {a_1^*}{a_2^*}
    \end{align*}
  \caption{Duality Involution}
  \label{fig:duality}
\end{figure}

\section{Semantics}
Next, we develop the basics of the model theory for \ohol{}. Being a
2-level type theory, \ohol{} has models that are a kind of
\emph{hyperdoctrine}. First, a model must include a model of
intuitionistic type theory to interpret the type-level context
$\Gamma$ and corresponding type/term judgments. Next, there should be
a functor from the category of semantic contexts and substitutions to
a category of models of ordered higher-order logic.

\subsection{Soundness and Completeness}

>>>>

\begin{abstract}
  We present a domain-specific type theory for constructions and
  proofs in category theory. The type theory axiomatizes notions of
  category, functor, profunctor and a generalized form of natural
  transformations. By restricting the type theory to an ordered linear
  restriction of predicate logic, we guarantee that all functions
  between categories are functorial, all relations are profunctorial,
  and all transformations are natural by construction, with no
  separate proof necessary. Important category theoretic proofs such
  as the Yoneda lemma and Co-yoneda lemma become simple type theoretic
  proofs about the relationship between unit, tensor and (ordered)
  function types, and can be seen to be ordered refinements of
  theorems in predicate logic.

  We develop a sound and complete notion of categorical model and give
  a model construction that shows that the theorems are valid in
  enriched and internal categories. This means that while the proofs
  in our type theory look like standard set-based arguments, the
  substructural syntactic restrictions ensure that the proofs carry
  over to enriched and internal settings as well.
\end{abstract}

@@

Category theory is a branch of mathematics that gives a systematic
theory of typed, ``higher-dimensional'' algebraic structures.
%
Originally developed for applications to homological algebra, it was
quickly discovered that these categorical structures were common in
logic and computer science, and that formal systems like logics, type
theories and programming languages can be modeled by sound and
complete semantic models given by notions of structured
category\cite{lawvere,goguen,lambek,moggi}, a phenomenon sometimes
referred to as the Curry-Howard-Lambek correspondence
\cite{curry-howard-lambek}.
%
Examples include the simply typed lambda calculus \cite{lambekscott},
computational lambda calculus \cite{moggi}, linear logic \cite{girard}
and dependent type theory\cite{not-sure}.
%
These seminal works have led to a fruitful field of research known as
categorical logic, where
new type theories are designed based directly on category-theoretic
semantics.
%
The slogan is that the syntax of a type theory should present an
initial object in its category of models, a category-theoretic
reformulation and generalization of the logical notion of soundness
and completeness.

% TODO: do we have citations for this?
While this research program has been quite successful,
category-theoretic techniques in computer science are sometimes
criticized for being overly complex, and there is difficulty in
acquisition of competence in category theory by computer scientists
who do not have a strong background in pure mathemantics.
%
While much of this can be explained by lack of, or lack of awareness
of, accessible material for computer scientists, in some ways category
theory is objectively complex.
%
In a traditional set-theoretic formulation, notions such as adjoint
functors and Kan extensions produce a proliferation of ``naturality''
and ``functoriality'' side-conditions that must be discharged.
%
This is compounded when we move from ordinary category theory to
enriched and internal category theory, where constructions must be
additionally proven continuous, monotone, etc, in addition to natural
or functorial.
%
These generalizations have long been recognized as being important in
programming language applications: domain-,metric- and
step-index-enriched categories have been used to model general
recursion and internal categories have been used to model
parametricity and gradual
typing\cite{order-enriched-cats,topos-of-trees,reflexive-graphs,double-cats-gradual-typing}.

@@

Fortunately, the tools of category theory itself can be employed to
simplify these constructions, specifically the tools of \emph{higher}
category theory.
%
Consider the analogous situation in differential calculus. When an
adept analyst writes down a function, they do not expand out the
$\epsilon-\delta$ definition of continuity and prove that it is
continuous. Rather, they are aware of certain \emph{syntactic
principles}, most importantly that composition of continuous functions
is continuous and that previously defined functions are known to be
continuous.
%
Similar principles apply to category theory itself: a composition of
functors is functorial, a composition of natural transformations is
natural, etc.
%
In the case of category theory, these principles can be formalized
using algebraic structures such as 2-categories, bicategories,
(virtual) double categories and pro-arrow equipments\cite{}.
%
This approach has the advantage of avoiding the need to prove
the side conditions of naturality or enrichment.
%
The downside is that these algebraic structures are quite complicated,
with a 2-dimensional diagrammatic language that can be quite beautiful
and elegant, but is also seems very far removed from the traditional
formulation of category theory that while complex is at least
described in terms of a concrete notion of sets or types.

In this work, we apply the technique of categorical logic to these
algebraic structures to design a formal logic, which we call
\emph{ordered predicate logic}, that can be soundly interpreted in
virtual equipments.
%
In ordered predicate logic, the sorts are interpreted as categories,
functions as functors, relations as profunctors and implications of
relations as natural transformations.
%
The type theoretic syntax allows us to write all of our constructions
as if they were ordinary functors and profunctors in set-based
category theory, but the proofs are valid in an arbitrary virtual
equipments and so are valid in enriched and internal category theory
as well.
%
This is analogous to the situation in intuitionistic type theory,
where the syntax is similar to that of informal set theory, but by
restricting to intuitionistic reasoning, the constructions can all be
interpreted in models where types are spaces and terms are continuous
functions ``for free''.

Ordered predicate logic is a kind of \emph{ordered},
\emph{constructive} variant of predicate logic over a unary type
theory.
%
It is \emph{ordered} in that our notion of predicate is that of a
binary relation $R(\alpha;\beta)$ where the $\alpha$ is thought of as
a contravariant dependence and the $\beta$ has a covariant dependence.
%
Proofs in the predicate logic $R_1, R_2,\ldots \vdash S$ are
restricted as in ordered linear logic \cite{lambek-calculus, ordered-linear-logic} in
that each input must be used exactly once and in the order they were
declared.
%
This order discipline ensures that every proof in the logic can be
interpreted as a \emph{natural} transformation.
%
The logic is \emph{constructive} in that we care about the identity of
proofs: they are natural transformations and there may be more than
one natural transformation of the same type.
%
If we ignore the identity of proofs, ordered predicate logic can be
viewed as a logic of preordered sets rather than categories.

The restriction to unary type theory and binary relations is fairly
strict: we cannot discuss important concepts such as functor
categories, opposite categories and arbitrary (co)-ends of
profunctors. However, the restriction is natural in that it
corresponds to a well-understood notion of model (virtual equipments)
and so we leave the extension to future work.
%
We show that a great deal of fundamental results in category theory
are already expressible in this fragment.

Due to the restriction to unary functors and binary profunctors, it is
difficult to state some theorems internally to the theory. For
instance, since we cannot express functor categories, we cannot
express any theorem that quantifies over functors purely in ordered
predicate logic, and instead only as a metatheorem. To more easily
formalize these metatheorems, we add a third layer to the type theory:
an ordinary dependent type theory in the style of Martin-L\"of type
theory. Then all of our ordered predicate logic judgments will also be
indexed by a context from this dependent type theory, and the type
theory includes types for categories, functors, profunctors and
natural transformations that allow us to formalize theorems the logic
is too restrictive to encode.

The dependent type theory layer also has an additional purpose: since
we can quantify over the structures in ordered predicate logic, we can
additionally encode \emph{signatures} for ordered predicate logic as
simply contexts of the dependent type theory. This allows for a clean
formalization of many categorical structures such as monoidal
categories, or strong monads.

While we emphasize the applications to enriched and internal category
theory in this work, there is potential for more direct application to
programming language semantics.
%
Ordinary predicate logic has been used to develop logics in which
logical relations proofs can be developed such as Abadi-Plotkin logic
for parametricity and LSLR for step-indexed logical relations proofs
\cite{abadi-plotkin,lslr}.
%
Many \emph{binary} logical relations are based on ordered structures,
such as logical relations for approximation in non-deterministic or
partial languages or logical relations for graduality in gradual
typing.
%
We conjecture that some extension of the ordered predicate logic
defined here will be useful for describing these relations as well.

\subsection{Basics of Profunctors}

While we assume the reader has some background knowledge of category
theory, we briefly define profunctors, which are not included in many
introductory texts. First, a category $\cat C$ has a collection of
objects and morphisms with identity and composition, and a functor $F
: \cat C \to \cat D$ is a function on objects and a function on
morphisms that respects source and target. It will be useful for
intuition to think of a category as a generalization of a preordered
set, which has a set of elements and a binary \emph{relation} on its
objects satisfying reflexivity and transitivity. It is fruitful to
think of a category as a \emph{proof-relevant preorder}, where
morphisms are thought of as proofs of ordering and the reflexivity and
transitivity constructions must additionally satisfy equations that
make them act like identity and composition. Then a functor is a
constructive version of a monotone function.

Given categories $\mathcal C$ and $\mathcal D$, a profunctor $R$ from
$\mathcal C$ to $\mathcal D$, written $R : \cat C \pto \cat D$ is a
functor $R : {\cat C}^{op} \times {\cat D} \to \Set$. We can think of
each set $R(c,d)$ as a set of ``heteromorphisms'' from $c$, an object
of $\cat C$ to $d$, an object of $\cat D$. Functoriality then defines
a notion of composition of heteromorphisms with homomorphisms in $\cat
C$ and $\cat D$. Thinking of categories as constructive preorders, a
profunctor is a constructive relation that is downward-closed in $\cat
C$ and upward-closed in $\cat D$, a kind of ``hetero-ordering''.
%
Given profunctors $R, S : \cat C \pto \cat D$, a homomorphism from
$R$ to $S$ is simply a natural transformation, which in the preordered
setting is simply an implication of relations.

The simplest example of a profunctor is for any category $\cat C$,
$\text{Hom}_{\cat C} : \cat C \pto \cat C$ is a profunctor. On
preorders this is just the ordering itself. The hom profunctor is the
unit for a composition of profunctors $R \odot S$ which is defined as
a \emph{co-end}. The heteromorphisms of $(R \odot S)(c,e)$ are triples
$(d \in \cat D, f \in R(c,d), g \in S(d,e))$ modulo the equivalence
relation generated by $(d,f,g) ~ (d', f', g')$ if there exists $h \in
D(d,d')$ such that $h\circ f = f'$ and $g'\circ h = g$. The
composition of profunctors is a generalization of the composition of
relations, and the hom profunctor is analogous in the same way to the
diagonal, or equality relation in that it is the identity for this
composition.
For any profunctor $R : \cat C \pto \cat D$ and functors $F : \cat
C' \to \cat C$ and $G : \cat D' \to \cat D$, by composition we can
define a profunctor $R(F,G)$, the ``restriction'' of $R$ along $F$ and
$G$.

Profunctors are not explicitly defined in most introductory texts on
category theory but arise implicitly in formulations of universal
properties such a representable functors and adjunctions.

@@

\item In \cref{sec:examples} we show how to use our syntax for
  synthetic category theory. Much of this is just rendering well-known
  2-categorical/double categorical constructions in our type theoretic
  syntax to demonstrate the ``naive''/synthetic character of proofs
  and constructions in SCT.

@@

\item In \cref{sec:semantics}, we develop some basic model theory for
  ordered predicate logic, including a sound and complete notion of
  categorical model and sound interpretation in virtual equipments
  modeling ordinary, enriched and internal category theory.

@@

@@

In Figure~\ref{fig:hol-cats} we give a table summarizing the
relationship between the judgments and connectives of higher-order
predicate logic with our ordered variant. Due to the incorporation of
variance, some unorderd concepts generalize to multiple different
ordered notions. For instance, covariant and contravariant presheaf
categories generalize the power set. Further, because we only have
binary relations rather than relations of arbitrary arity, we have
only restricted forms of universal and existential quantification
which come combined with implications and conjunctions.

@@

\subsection{Duality Principle}
One common source of category theoretic arguments is the duality
principle: since the opposite of any category is also a category, any
statement that quantifies over all categories is equally true of their
opposites. Furthermore in a formal language of categories, any
statement can be dualized to a statement where domains and codomains
are swapped and composition is flipped.
%
A similar duality principle holds in SCT.
%
The basic idea is to reverse the orientation of all \emph{set}
expressions by swapping which variable is covariant and which is
contravariant.
%
This has the effect of reversing the order of the context $\Phi$.
%
We can define this as a translation $\cdot^* : SCT \to SCT$ defined on
syntactic forms of SCT in Figure~\ref{fig:duality}.
\begin{figure}
    \begin{align*}
      (\harr {\cat C} {\cat D})^* &= \harr {\cat D^*} {\cat C^*}\\
      (\alpha:\cat C)^* &= \alpha: \cat C^*\\
      (\alpha:\cat C; \beta:\cat D)^* &= \beta:\cat D^*; \alpha:\cat C^*\\
      (\Phi,x:R,\Psi)^* &= \Psi^*,x:R^*, \Phi^*\\
      (\homrallXYtoZ \alpha P Q)^* &= \homlallXYtoZ \alpha {P^*} {Q^*}\\
      (\homlallXYtoZ \alpha P Q)^* &= \homrallXYtoZ \alpha {P^*} {Q^*}\\
      (\homrlambdaXatYdotZ x \alpha s)^* &= \homllambdaXatYdotZ x \alpha s^*\\
      (\homrappXtoYatZ s t a)^* &= \homlappXtoYatZ {s^*} {t^*} {a^*}\\
      (\tensorexistsXwithYandZ \beta P Q)^* &=
      \tensorexistsXwithYandZ \beta {Q^*}{P^*}\\
      (\tensorintroatXwithYandZ b s t)^* &= \tensorintroatXwithYandZ {b^*} {t^*} {s^*}\\
      (\punitinXfromYtoZ {\cat C} {a_1}{a_2})^* &= \punitinXfromYtoZ {\cat C^*} {a_2^*}{a_1^*}
    \end{align*}
  \caption{Duality Involution}
  \label{fig:duality}
\end{figure}

Next, we develop the basics of the model theory for \ohol{}.
%
First, we demonstrate that a hyperdoctrine of virtual double
categories with split restrictions provides a sound and complete
notion of categorical model.
%
Then we instantiate this general notion of model to show that the
\ohol{} can be interpreted in ordinary category theory as well as
enriched and internal category theory, and more trivially, in terms of
sets and relations or preordered sets and ordered relations.

\subsection{Categorical Models}

First, we can model the judgmental structure of the unary type theory
and predicate logic in certain\emph{virtual double categories}
\cite{leinster,crutwell-shulman}. We briefly recount the structure
present in a virtual double category, but see \citet{crutwell-shulman}
for a precise definition.
The following definition is
equivalent to the formulation as a ``normalized $T$-monoid'' in
\citet{crutwell-shulman}, and we direct the reader there for the full
description.
%

\begin{definition}[Virtual Double Category]
  A virtual double category $\mathcal V$ consists of
  \begin{enumerate}
  \item A category $V_o$ of ``objects and vertical arrows''
  \item A set ${\mathcal V}_h$ of ``horizontal arrows'' with source
    and target functions $s,t : {\mathcal V}_h \to {\mathcal V_o}^2$
  \item Sets of 2-cells of the following form:
% https://q.uiver.app/?q=WzAsNixbMCwwLCJDXzAiXSxbMCwyLCJEXzAiXSxbMiwwLCJDX3tufSJdLFsxLDAsIlxcY2RvdHMiXSxbMiwyLCJEXzEiXSxbMSwxLCJcXHBoaSJdLFszLDIsIlJfbiIsMCx7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImJhcnJlZCJ9fX1dLFswLDEsIkYiLDJdLFsyLDQsIkciXSxbMSw0LCJTIiwyLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn19fV0sWzAsMywiUl8wIiwwLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn19fV1d
\[\begin{tikzcd}
	{C_0} & \cdots & {C_{n}} \\
	& \phi \\
	{D_0} && {D_1}
	\arrow["{R_n}", "\shortmid"{marking}, from=1-2, to=1-3]
	\arrow["F"', from=1-1, to=3-1]
	\arrow["G", from=1-3, to=3-3]
	\arrow["S"', "\shortmid"{marking}, from=3-1, to=3-3]
	\arrow["{R_0}", "\shortmid"{marking}, from=1-1, to=1-2]
\end{tikzcd}\]
  \end{enumerate}
  We say a virtual double category has \emph{split restrictions} when
  for any horizontal arrow $R : C \pto D$ and vertical arrows $f : C'
  \to C$ and $g : D' \to D$ there is a chosen profunctor $R(f,g) : C'
  \pto D'$ with a cartesian 2-cell to $R$ framed by $f,g$ and these
  chosen cartesian lifts are functorial in $f,g$.  A functor of
  virtual double categories preserves split restrictions if it
  preserves these specified cartesian lifts.  
\end{definition}

Each component of this definition has a direct correspondence to a
syntactic structure in \ohol{}. The category $\mathcal V_o$ models the
category and functor judgments. The set $\mathcal V_h$ models the
profunctor judgment. A composable string $R_0 \cdots R_n$ models the
profunctor contexts. The 2-cells correspond to the natural
transformation judgment where we have taken the restriction $S(F,G)$
of the codomain.

To model the dependent type theory and fact that the \ohol{} judgments
are all under a dependent type theory context, we use a variation on
Lawvere's notion of \emph{hyperdoctrine} for modeling predicate logic:
\begin{definition}[Hyperdoctrine of Virtual Double Categories with Restrictions]
  A hyperdoctrine of virtual double categories with restrictions is a
  pair of a category with families $\mathcal C$ and a functor $L :
  \mathcal C^o \to \text{vDbl}_r$ to the category of virtual double
  categories with split restrictions and restriction-preserving
  functors.
\end{definition}
Categories with families $\mathcal C$ model dependent type theory
\citep{cwf} and for each semantic context $\Gamma$, $L(\Gamma)$ models
the \ohol{} judgments in context $\Gamma$, with the functoriality
modeling the fact that all of these judgments admit a well-behaved
action of substitution.

\begin{construction}[Soundness and Completeness]
  For any choice of connectives, the syntax of parameterized \ohol{}
  defines a hyperdoctrine of VDCrs that is an initial object in the
  category of hyperdoctrines of VDCrs with chosen universal
  constructions and functors preserving those universal constructions.
\end{construction}
\begin{proof}
  Most of the details of the proof are straightforward. In the
  appendix we demonstrate the syntactic elements of the proof which
  are that the syntax itself is a hyperdoctrine of VDCrs, i.e., that
  the notions of substitution are appropriately functorial.
\end{proof}
